Biogeography and Environmental Conditions Shape Phage and Bacteria Interaction Networks Across the Healthy Human Microbiome

### 2017-05-08 doc/manuscript.Rmd
Updates.

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. Such approaches are not able to capture how these communities interact to maintain stability/diversity, co-evolve, and share genetic information (e.g. horizontal gene transfer <!--X as this isn't the only way to share information, use eg instaed of ie. Also, is there work that demonstrates how these communities interact to maintain stability? By stability, do we mean maintain diversity? What are the other outcomes of interactions? e.g., (antagonistic) co-evolution, etc. I know these aren't elements to fit into an Abstract, so staying tuned as to whether they appear later, as well-->). We developed and implemented a network-based analytical approach to describe phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. This study quantified and contrasted the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to investigate the role of virus-microbe ecological networks in advancing our understanding of conditions, such as disease states.

@@

We investigated human bacterial-phage networks by utilizing three published microbiome datasets which contained paired virus and total metagenomic sequence sets, which primarily consisted of bacteria [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models [@Edwards:2015iz, @Roux:2016cc] by its inclusion of protein interaction data as well as the inclusion of both negative and positive interactions. Through this approach, we were able to provide a basic understanding of the network dynamics associated with healthy human phage and bacterial communities. By building and utilizing a microbiome-virome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we empower future studies to begin investigating how these community structures are influenced by disease state and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, which explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. These datasets were selected because they included virome samples subjected to virus-like particle (VLP) purification. They employed combinations of filtration, chloroform/DNase treatment, and Cesium Chloride treatments to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria and other microbiome members (e.g. fungi, archaea, *Demodex* mites, etc) were not considered in this initial study.

Here, the bacterial and viral sequences from these studies were quality filtered and assembled into contigs. We generated clusters of related bacterial and phage contigs based on their k-mer frequencies and co-abundance patterns using the CONCOCT algorithm **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. Contig lengths within clusters ranged between $10_{3}$ and $10_{5.5}$, confirming the majority of contigs represented genome fragments, rather than complete genomes  **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. We also confirmed that the majority of bacterial clusters were indeed identified as bacteria (CITE FIGURE HERE). We interpreted these clusters as operational units of related bacteria and phage genomes, which we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht].

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacterial OGUs using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. This training set contained 43 diverse bacterial species and 30 diverse phage strains, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because we used DNA sequencing studies, RNA phages were not considered **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (a phage infects a bacterium) and negative relationships (a phage does not infect a bacterium). This allowed us to validate the false positive and false negative rates associated with our candidate models, thereby building upon previous work that only considered positive relationships [@Edwards:2015iz].

@@

# Discussion
<!-- Much of the discussion repeats the statements in the results, it felt a bit redundant. At times, the discussion even repeated itself (you'll see I deleted most of the last paragraph). There are probably ways to be less repetitive and more provocative or to bring in discussion of more recent work? Or to reduce it even more by removing redundancies. Alternatively, could remove the redundancies in the results...as in only state the results, and not reference previous work. Then could discuss how findings relate to previous work only in the discussion, rather than the results...which does happen at times in current form.-->
Foundational microbiome work has provided a baseline understanding of the human microbiome by characterizing bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Here, we offer an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how and why microbiome networks differ in disease states. We developed and implemented a network-based analytical approach to evaluate the basic properties of the human microbiome through bacteria and phage relationships, instead of membership or diversity alone. This enabled the application of network theory to provide a new perspective on complex ecological communities. We utilized metrics of connectivity to model the extent to which communities of bacteria and phages interact, through mechanisms including horizontal gene transfer, modulated bacterial gene expression, and alterations in abundance.

Just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were conserved within individuals over time. Yet, gut network structure was not conserved among family members. These findings suggested that microbiome network properties, including stability and the potential for horizontal gene transfer <!-- and lots of other things related to host and virus eco and evo processes -->, were personal and may be impacted by personal factors ranging from the body's immune system to environmental conditions, such as climate and diet <!-- though I'd argue that the immune system is a component/control on the environmental condition of the microbial communities -->. The ability of environmental conditions to shape gut and skin microbiome network structure was supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, our findings suggested that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability for microbes to directly influence one other. We supported this finding with the observation that obesity was associated with decreased network connectivity (closeness of bacterial and phage vertices). Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their interactions. Further work will be required to characterize these relationships with a larger cohort.

In addition to diet, the skin microbiome network structure varied by skin environment. Network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure, the exposed and intermittently exposed sites, were less connected and therefore were predicted to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Together these data suggested body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

While these findings take us an important step closer to understanding the microbiome through interspecies relationships, there are caveats to the approach that should be noted. First, although our infection classification model is advantageous over existing models, we recognize that, like most classification models, ours is only as good as its training set. Large-scale experimental screens for phage and bacteria infectious interactions that report high-confidence negative interactions (i.e., no infection) will provide more robust model training and improved model performance. Just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will be integrated into this model to further improve performance. Second, although our analyses offer an informative proof of concept, this work was done retrospectively and relied on published research from as long as seven years ago. These archived datasets were limited by the technology and costs of the time, meaning the datasets were suboptimally powered for the statistical analysis we strive for today. Despite these limitations, we were able to present initial conclusions. Follow-up studies will validate our findings and inform the design and interpretation of future studies.

Our work takes a first step towards defining bacteria-virus interaction profiles as a characteristic of human-associated microbial communities. We provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be applied therapeutically.

>>>>

# Abstract
Viruses and bacteria are critical components of the human microbiome and play important roles in health and disease. Most previous work has relied on studying bacteria and phages in isolation, reducing them to two separate communities. Such approaches are not able to capture how these communities interact to maintain stability/diversity, co-evolve, and share genetic information (e.g. horizontal gene transfer). We developed and implemented a network-based analytical approach to describe phage-bacteria network diversity throughout the human body. We accomplished this by building a machine learning algorithm to predict which phages infected which bacteria in a given microbiome. This algorithm was applied to paired viral and bacterial metagenomic sequence sets from three previously published human cohorts. We organized the predicted interactions into networks that allowed us to evaluate phage-bacteria connectedness across the human body. We found that gut and skin network structures were person-specific and were not conserved among cohabitating family members. High-fat diets and obesity were associated with less connected networks. There were significant differences in network structure between skin sites, with those exposed to the external environment being less connected and more prone to instability. This study quantified and contrasted the diversity of virome-microbiome networks across the human body and illustrated how environmental factors may influence phage-bacteria interactive dynamics. This work provides a baseline for future studies to investigate the role of virus-microbe ecological networks in advancing our understanding of conditions, such as disease states.

@@

We investigated human bacterial-phage networks by utilizing three published microbiome datasets that contained paired virus and total metagenomic sequence sets, which primarily consisted of bacteria [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei]. Our approach built off of previous large-scale phage-bacteria microbiome network analyses by inferring interactions using metagenomic datasets, instead of using culture-based techniques [@Flores:2013hc]. This metagenomic interaction inference model was powered beyond previous models [@Edwards:2015iz, @Roux:2016cc] by its inclusion of protein interaction data as well as the inclusion of both negative and positive interactions. Through this approach, we were able to provide a basic understanding of the network dynamics associated with healthy human phage and bacterial communities. By building and utilizing a microbiome-virome network, we found that different people, body sites, and anatomical locations not only support distinct microbiome membership and diversity [@Hannigan:2015fz; @Minot:2011ez; @Reyes:2010cwa; @Turnbaugh:2009ei; @Grice:2009eea; @Findley:2013jf; @Costello:2009im, @Consortium:2012iz], but also support communities with distinct communication structures and propensities toward community instability. Through an improved understanding of the healthy state of network structures across the human body, we empower future studies to begin investigating how these community structures are influenced by disease state and their overall impact on human health.

@@

## Cohort Curation and Sample Processing
We studied the differences in virus-bacteria interaction networks across healthy human bodies by leveraging previously published sequence sets containing purified virome samples paired with bacterial metagenomes from whole metagenomic shotgun sequences. Our study contained three datasets, which explored the impact of diet on the healthy human gut virome [@Minot:2011ez], the impact of anatomical location on the healthy human skin virome [@Hannigan:2015fz], and the viromes of monozygotic twins and their mothers [@Reyes:2010cwa; @Turnbaugh:2009ei]. These datasets were selected because they included virome samples subjected to virus-like particle (VLP) purification. They employed combinations of filtration, chloroform/DNase treatment, and Cesium Chloride treatments to eliminate organismal DNA and thereby allow for direct assessment of both the extracellular and fully-assembled intracellular virome **(Supplemental Figure \ref{SequenceStats} A-B)** [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. While the whole metagenomic shotgun sequence samples were not subjected to purification, they primarily consisted of bacteria and had minimal representation from other groups [@Minot:2011ez, @Hannigan:2015fz, @Reyes:2010cwa; @Turnbaugh:2009ei]. Other, minimally represented microbiome members (e.g. fungi, archaea, eukaryotes including *Demodex* mites, etc) were not considered in this study.

Here, the bacterial and viral sequences from these studies were quality filtered and assembled into contigs. We generated clusters of related bacterial and phage contigs based on their k-mer frequencies and co-abundance patterns using the CONCOCT algorithm **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. Contig lengths within clusters ranged between $10^{3}$ and $10^{5.5}$, confirming the majority of contigs represented genome fragments, rather than complete genomes  **(Supplemental Figure \ref{ContigStats} - \ref{ClusterStats})**. We also confirmed that the majority of bacterial clusters were indeed identified as bacteria (CITE FIGURE HERE), confirming the previous reports [@Minot:2011ez, @Hannigan:2015fz, @Turnbaugh:2009ei]. We interpreted these clusters as operational units of related bacteria and phage genomes, which we defined as operational genomic units (OGUs). These OGUs are conceptually similar to the operational taxonomic unit (OTU) and operational protein family (OPF) definitions used for grouping highly similar 16S rRNA gene and open reading frame sequences, respectively [@Schloss:2008ht].

## Establishing a Model of Phage-Bacteria Interactions
We predicted which phage OGUs infected which bacterial OGUs using a random forest model trained on experimentally validated infectious relationships from six previous publications [@Jensen:1998vh;@Malki:2015tm;@Schwarzer:2012ez;@Kim:2012dh;@Matsuzaki:1992gw;@Edwards:2015iz]. Only bacteria and phages were used in the model. The training set contained 43 diverse bacterial species and 30 diverse phage strains, with both broad and specific infectious ranges **(Supplemental Figure \ref{ValidationOverview} A - B)**. Phages with linear and circular genomes, as well as ssDNA and dsDNA genomes, were included in the analysis. Because we used DNA sequencing studies, RNA phages were not considered **(Supplemental Figure \ref{ValidationOverview} C-D)**. This training set included both positive relationships (a phage infects a bacterium) and negative relationships (a phage does not infect a bacterium). This allowed us to validate the false positive and false negative rates associated with our candidate models, thereby building upon previous work that only considered positive relationships [@Edwards:2015iz].

@@

# Discussion
Foundational microbiome work has provided a baseline understanding of the human microbiome by characterizing bacteria and viral diversity across the human body, as well as other environments [@Grice:2009eea; @Findley:2013jf; @Hannigan:2015fz; @Costello:2009im, @Consortium:2012iz; @Schloss:2005hz; @Minot:2011ez]. Here, we offer an initial understanding of how phage-bacteria networks differ throughout the human body, so as to provide a baseline for future studies of how and why microbiome networks differ in disease states. We developed and implemented a network-based analytical approach to evaluate the basic properties of the human microbiome through bacteria and phage relationships, instead of membership or diversity alone. This enabled the application of network theory to provide a new perspective on complex ecological communities. We utilized metrics of connectivity to model the extent to which communities of bacteria and phages interact, through mechanisms including horizontal gene transfer, modulated bacterial gene expression, and alterations in abundance.

Just as gut microbiome and virome composition and diversity are conserved in individuals [@Hannigan:2015fz; @Grice:2009eea; @Findley:2013jf; @Minot:2013ih], gut and skin microbiome network structures were conserved within individuals over time. Gut network structure was not conserved among family members. These findings suggested that microbiome network properties, including stability, the potential for horizontal gene transfer, co-evolution, etc, were personal and may be impacted by personal factors ranging from the body's immune system to external environmental conditions, such as climate and diet. The ability of environmental conditions to shape gut and skin microbiome network structure was further supported by our finding that diet and skin location were associated with altered network structures.

We found evidence that diet was sufficient to alter gut microbiome network connectivity. Although our sample size was small, our findings provided evidence that high-fat diets were less connected than low-fat diets. This suggested that high-fat diets may lead to less stable communities with a decreased ability for microbes to directly influence one another. We supported this finding with the observation that obesity may have been associated with decreased network connectivity. Together these findings suggest the food we eat may not only impact which microbes colonize our guts, but may also impact their interactions. Further work will certainly be required to characterize these relationships with a larger cohort.

In addition to diet, the skin microbiome network structure varied by skin environment. Network structure differed between environmentally exposed and occluded skin sites. The sites under greater environmental fluctuation and exposure (the exposed and intermittently exposed sites) were less connected and therefore were predicted to have a higher propensity for instability. Likewise, intermittently moist sites demonstrated less connectedness than the more stable moist and sebaceous sites. Together these data suggested that body sites under greater degrees of fluctuation harbored less connected, potentially less stable microbiomes.

While these findings take us an important step closer to understanding the microbiome through interspecies relationships, there are caveats to the approach that should be noted. First, although our infection classification model is advantageous over existing models, we recognize that, like most classification models, ours is only as good as its training set. Large-scale experimental screens for phage and bacteria infectious interactions that report high-confidence negative interactions (i.e., no infection) will provide more robust model training and improved model performance. Furthermore, just as we have improved on previous modeling efforts, we expect that new and creative scoring metrics will be integrated into this model to improve performance.

Second, although our analyses offer an informative proof of concept, this work was done retrospectively and relied on published research from as long as seven years ago. These archived datasets were limited by the technology and costs of the time, meaning the datasets were suboptimally powered for the statistical analysis we strive for today. Despite these limitations, we were able to present initial conclusions. Follow-up studies will validate our findings and inform the design and interpretation of future studies.

It is important to note that the microbiome consists of bacteria and viruses, as well as fungi, archaea, eukaryotes such as *Demodex* mites, etc. These microbiome components also interact with human immune cells and other non-microbial community members. Our study only represents an initial step toward network analysis, and future work will include more components of the microbiome to gain a more complete understanding of the community through its interactions. In addition to these microbiome components, additional relationship types will also be added so as to provide a more robust network for analysis.

Together, our work takes a first step towards defining bacteria-virus interaction profiles as a characteristic of human-associated microbial communities. By focusing on relationships between bacterial and viral communities, we can begin studying them as interacting cohorts they are, instead of isolated entities. This work will also provide a foundation for future studies to begin evaluating how microbiome network dynamics change in disease states, and how the information can be applied therapeutically.
